Møil: A web-based administration tool for mail servers backed by relational databases

### 2014-12-10 doc/study/study.tex
Optional line numbers for reviews.
Include date in title.

@@

	\title{Møil: A web-based administration tool for mail servers backed by relational databases}
	\author{Henning Müller $<$\href{mailto:henning@orgizm.net}{henning@orgizm.net}$>$}
	\date{}

@@

	\section{Abstract}
		This document compares existing administration user interface solutions
		for mail servers running \emph{postfix} and backed by relational
		databases and introduces yet another one, which is called \emph{Møil}
		and is written in Ruby utilizing the
		\ac{Rails}\footnote{\url{http://rubyonrails.org}} framework. It brings
		the handy possibility of managing the database schema with migrations,
		a lot of beautiful, responsive \acs{CRUD} and some other nice features
		(see \emph{Møil} \textref{sec:moeil:features}). The code is available
		from GitHub\footnote{\url{https://github.com/nning/moeil}} licensed
		under AGPLv3 \cite{agpl}.

@@

		For mail system setups (meaning one or more servers running Mail
		Submission, Delivery and Transfer Agent software and providing a
		Message Store for users \cite{mail-architecture}), storing data about
		which domains and E-Mail addresses the system is responsible for
		(further called meta data) in relational databases (as opposed to plain
		files for example) is beneficial for several reasons.

@@

		In bigger setups, it offers the possibility for scaling to more than
		one node (horizontal scalability) to cope with high mail and user
		loads: Several servers handling incoming mail or users reading their
		mail act as readers on one or more database servers. The performance of
		the frequent look-up of domains, mailboxes and aliases (forwardings)
		for receiving mail or user logins can be increased. The actual
		accommodation of E-Mails -- another big problem with horizontal scaling
		-- can not be considered here.

@@

		\acp{DBMS} for relational databases usually provide a query interface
		with the \ac{SQL} \cite{sql} for editing the data sets. It is widely
		supported by programming languages and equally widely spoken among
		programmers, which makes it valuable for integration of several
		services dealing with this mail system meta data. For example this can
		be used to integrate accounting solutions for charging purposes into
		mail systems.

@@

		\subsection{Criteria}
		\label{sec:evaluation:criteria}
			Not only for the development of \emph{Møil}, some of the criteria
			arose from shortcomings of \emph{postfix.admin} (see page
			\pageref{sec:contestants:postfix.admin}), which is a quite widely
			deployed solution. The criteria will be classified into four
			categories (Data model, features, security \& robustness and user
			interface) and numbered for easier reference. An example for the
			textual distinguation is the criterion \crit{normalized database
			schema}{1.1} as it is written here. There is some meta-data like
			website, programming language, last update in VCS and number of
			commits, which is not numbered and compared. The used programming
			language would be a neutral attribute but gets a criterion, because
			before the development of \emph{Møil}, a decision was made against
			deploying further PHP applications to accomplish a more
			maintainable and secure web hosting
			environment\footnote{\urlPHPBashing}.

@@

			\subsubsection{Features (2)}
				There are features, which no mail server administration user
				interface could spare and yet they will be listed here, because
				it is important to analyze, how these are implemented and how
				the implementations differ by solution. The \crit{basic
				domain, mailbox and alias management}{2.1} is one of them. The
				\crit{password changing ability for users}{2.2} is a feature,
				which is also quite basic and the first one legitimating
				letting users access the administration front-end. Some
				solutions also integrate into popular self-hosted web mailer
				software for password changing. When users are allowed to
				access the administration front-end, \crit{access control}{2.3}
				or authorization is needed to selectively restrict the access
				to resources. It could also be possible to delegate
				administrative access on domains or resources in general to
				certain users.

				More mature utilities also bring useful non-basic features as
				an \crit{audit trail}{2.4.1} for example, which can be used to
				log any changes to the database with time and transacting
				account. It is also possible, the audit trail provides an undo
				functionality. In some environments users demand \crit{vacation
				mailing}{2.4.2} to automatically answer any incoming mail with
				a pre-defined text when they are vacant. Some solutions serve
				users with the possibility to activate or deactivate it and
				edit the pretext. Automatic answers can also be solved more
				generic with \crit{Sieve}{2.4.3} \cite{sieve}, which is a
				server-side mail filter interface. The protocol for users to
				edit the server-side filters is called \emph{ManageSieve}
				\cite{managesieve}. Another common demand of users of a mail
				system is a \crit{spam front-end}{2.4.4} to review mails, which
				got quarantined by spam or virus checking software. (Although
				in my personal experience this false negative classification
				does occur almost never with a setup common in the Open Source
				world consisting of \emph{amavisd-new}, \emph{spamassassin} and
				\emph{clamav}.)

			\subsubsection{Security \& robustness (3)}
				For now, only three criteria in this category shall be
				considered, because further security analysis would require an
				in-depth source code review. The \crit{password hashing}{3.1}
				availability and algorithm strength is important against
				unauthorized database access through potential security holes
				in the application. When the database is maliciously read, the
				value of the passwords to the attacker is inversely
				proportional to the time it costs, to crack the password
				hashing. \texttt{MD5} and \texttt{SHA1} based methods are
				considered unsecure, \texttt{SHA256}, \texttt{SHA384},
				\texttt{SHA512} and \texttt{bcrypt} are
				recommended\footnote{\url{http://www.keylength.com/en/8/}}\textsuperscript{,}\footnote{\url{http://codahale.com/how-to-safely-store-a-password}}.
				\crit{Test coverage}{3.2} is important for this evaluation and
				for security purposes, because it lowers the probability of
				bugs per lines of code by trend. As \crit{Containability}{3.3},
				the modular independence (e.g. independence of the web
				interface from the Message Store file system) from a security
				perspective is perceived.

@@

		\subsection{The contestants}
			For introduction of "the contestants", the most important
			attributes and specialties will be listed for each. A more detailed
			comparison of the features can be found under
			\textref{sec:evaluation:overview}.

@@

				\noindent
				Of the compared tools, \emph{modoboa} seems to be one of the
				most mature ones. It uses the Python Django framework and
				therefore includes facilities like schema migrations and an
				\ac{ORM}. It passes almost every criterion; just when it comes
				to the \crit{audit trail}{2.4.1}, there is no undo support.
				Also the user interface on a mobile phone is not fully
				elaborated and has some minor flaws.

@@

				\noindent
				Further information on \emph{Møil} is to be found in the
				section \textref{sec:moeil}.

@@

				\noindent
				For many of the solutions, \emph{postfix.admin} seems to be an
				inspiration and also the impulse for a new implementation. Many
				solutions aim to overcome its shortcomings, but many also fail
				to live up to that task (which in this case means every PHP
				tool). \emph{postfix.admin} uses a non-normalized \cite{dbnorm}
				database schema, which could spare a JOIN in some database
				queries but is very prone to inconsistencies. \crit{Schema
				migrations}{1.4} are not present through a database abstraction
				library or an \ac{ORM}, but manually implemented rather
				minimalist. There is one big file (\texttt{upgrade.php})
				containing a pile of functions like
				\texttt{upgrade\_<n>\_<db>}, which contain database dependent
				raw \ac{SQL} queries to perform the transformations on the
				schema. There is no \crit{systematic database
				independence}{1.5} or an \ac{ORM}, just "manual" support for
				MySQL and PostgreSQL. Oddly, for every mailbox there is an
				alias entry created (even when the mail is not forwarded).
				\emph{postfix.admin} has support for an \crit{audit
				trail}{2.4.1} but without undo. The default \crit{password
				hashing}{3.1} algorithm is \texttt{md5-crypt}. It is
				\crit{containable}{3.3} by default but does not support the
				movement of mail data, then, which would require the direct
				access to the Mail Store.

@@

				\item[\rm Responsive user interface]\ \\
					\emph{Møil} adapts to the environment, it is needed in. The
					web user interface is responsive and therefore supports
					both classic desktop view ports and mobile devices like
					phones or tablets. For friends of the command line, some
					commands to manage the most important resources are
					provided, too.

@@

				\item[\rm Search]\ \\
					A search for any resources is provided, which is optionally
					based on \ac{SQL} or
					\emph{Elasticsearch}\footnote{\url{http://www.elasticsearch.org}}
					for really huge databases (if installed and configured to
					be used).
			\end{description}

@@

			The application model is put into graphs in the following image. It
			shows the model classes and their relations among each other. (The
			graph is generated, so unfortunately (for now) the actual
			attributes saving a relation and the cardinalities are missing.)

@@

			\subsubsection{Mailboxes}
				A \texttt{Mailbox} is connected to its \texttt{Domain} and
				possibly also to one \texttt{Relocation} and has the attributes
				\texttt{username} (which is the part in front of the
				domain in the E-Mail address), \texttt{encrypted\_password}
				(containing the password hash), \texttt{name} (for a given name
				for example), \texttt{mail\_location} (which is empty by
				default but can be used to move the users mail data location),
				\texttt{quota}, \texttt{active}, \texttt{admin} and timestamp
				attributes for saving the date of creation or the last update.
				As a password hashing algorithm, currently only
				\texttt{sha512-crypt} is supported.

			\subsubsection{Aliases}
				\texttt{Aliases} represent mail forwardings and are also
				connected to their affiliated \texttt{Domain}. They have the
				attributes \texttt{username}, \texttt{goto} (which is the
				target of the forwarding and can contain one or more internal
				or external E-Mail addresses seperated by commas),
				\texttt{active}, \texttt{description} and timestamps of
				creation and last update.

@@

			\begin{description}
				\item[\rm Better \ac{CLI}]\ \\
					Currently, Møil has seperate \ac{CLI} tools for the most
					important tasks like managing domains and mailboxes and
					importing data from an \emph{postfix.admin} installation.
					This can be expanded into a single \ac{CLI} utility with
					better support for the features, the web user interface
					offers.

				\item[\rm Spam and virus check integration]\ \\
					Some compared solutions integrate \texttt{amavisd-new} by
					shared database tables. After this being clear, there is
					not much left to be solved conceptually for a spam and
					virus check integration and this becomes a question of time.

				\item[\rm Support for server-side filters]\ \\
					The support for server-side filters in Møil was already
					examined and begun but at first, a Ruby library for the
					ManageSieve protocol has to be implemented, because the
					only one existing is not maintained anymore and became
					disfunctional.
			\end{description}

>>>>

\IfFileExists{review.tex}{
	\include{review}
}

@@

	\title{Møil: A web-based administration tool for mail servers backed by relational databases}
	\author{Henning Müller $<$\href{mailto:henning@orgizm.net}{henning@orgizm.net}$>$}
	\date{\today}

@@

	\section{Abstract}
		This document compares existing administration user interface solutions
		for mail servers running \emph{postfix} and \emph{dovecot} backed by
		relational databases and introduces yet another one, called \emph{Møil}
		and written in Ruby utilizing the
		\ac{Rails}\footnote{\url{http://rubyonrails.org}} framework. It
		provides an usable, responsive \ac{UI}, database independence and ACLs.
		More features are listed in the \textsecref{sec:moeil:features}. The
		code is available from
		GitHub\footnote{\url{https://github.com/nning/moeil}} licensed under
		AGPLv3 \cite{agpl}.

@@

		For mail system setups (meaning one or more servers running Mail
		Submission, Delivery and Transfer Agent software and providing a
		Message Store for users \cite{mail-architecture}), storing data about
		which domains and email addresses the system is responsible for
		(further called meta data) in relational databases (as opposed to plain
		files for example) is beneficial for several reasons.

@@

		In bigger setups, it offers the possibility for scaling to more than
		one node (horizontal scalability) to cope with high mail and user
		loads: Several servers handling incoming mail or users reading their
		mail act as readers on one or more database servers. The performance of
		the frequent look-up of domains, mailboxes and aliases (forwardings)
		for receiving mail or user logins can be increased, because it offers
		more parallel read capacity. The actual accommodation of emails --
		another big problem with horizontal scaling	-- can not be covered
		here.

@@

		\acp{DBMS} for relational databases usually provide a query interface
		based on the \ac{SQL} \cite{sql} for editing the data sets. It is widely
		supported by programming languages and equally widely spoken among
		programmers, which makes it valuable for integration of several
		services dealing with this mail system meta data. For example this can
		be used to integrate accounting solutions for charging purposes into
		mail systems using \ac{SQL} as consistent interface to the application
		data (regarded as an \acs{API}).

@@

		\subsection{Criteria}
		\label{sec:evaluation:criteria}
			Some of the criteria arose from shortcomings of
			\emph{postfix.admin} (see page
			\pageref{sec:contestants:postfix.admin}), which is a quite widely
			deployed solution -- not only for the development of \emph{Møil}.
			The criteria will be classified into four categories (Data model,
			features, security \& robustness and user interface) and numbered
			for easier reference. An example for the textual distinguation is
			the criterion \crit{normalized database schema}{1.1} as it is
			written here: A short description in \textbf{bold} font succeeded
			by a sequential number in parenthesis. There is some meta-data like
			website, programming language, last update in VCS and number of
			commits, which is not numbered and compared. The used programming
			language would be a neutral attribute but gets a criterion, because
			before the development of \emph{Møil}, a decision was made against
			deploying further PHP applications to accomplish a more
			maintainable and secure web hosting
			environment\footnote{\urlPHPBashing}.

@@

			\subsubsection{Features (2)}
				There are features, which no mail server administration user
				interface could spare and yet they will be listed here, because
				it is important to analyze how these are implemented and how
				the implementations differ by solution. The \crit{basic
				domain, mailbox and alias management}{2.1} is one of them. The
				\crit{password changing ability for users}{2.2} is a feature,
				which is also quite basic and the first one legitimating
				letting users access the administration front-end. Some
				solutions also integrate into popular self-hosted web mailer
				software for password changing. When users are allowed to
				access the administration front-end, \crit{access control}{2.3}
				or authorization is needed to selectively restrict the access
				to resources. It could also be possible to delegate
				administrative access on domains or resources in general to
				certain users.

				More mature utilities also bring useful non-basic features as
				an \crit{audit trail}{2.4.1} for example, which can be used to
				log any changes to the database with time and transacting
				account. It is also possible the audit trail provides an undo
				functionality. In some environments users demand \crit{vacation
				mailing}{2.4.2} to automatically answer any incoming mail with
				a pre-defined text when they are vacant. Some solutions serve
				users with the possibility to activate or deactivate it and
				edit the pretext. Automatic answers can also be solved more
				generic with \crit{Sieve}{2.4.3} \cite{sieve}, which is a
				server-side mail filter interface. The protocol for users to
				edit the server-side filters is called \emph{ManageSieve}
				\cite{managesieve}. Another common demand of users of a mail
				system is a \crit{spam front-end}{2.4.4} to review mails, which
				got quarantined by spam or virus checking software. (Although
				in my personal experience this false negative classification
				does occur almost never with a setup common in the Open Source
				world consisting of \emph{amavisd-new}, \emph{spamassassin} and
				\emph{clamav}.)

			\subsubsection{Security \& robustness (3)}
				For now, only three criteria in this category shall be
				considered, because further security analysis would require an
				in-depth source code review. The \crit{password hashing}{3.1}
				availability and algorithm strength is important against
				unauthorized database access through potential security holes
				in the application. When the database is maliciously read, the
				value of the passwords to the attacker is inversely
				proportional to the time it costs to crack the password
				hashing. \texttt{MD5} and \texttt{SHA1} based methods are
				considered unsecure, \texttt{SHA256}, \texttt{SHA384},
				\texttt{SHA512} and \texttt{bcrypt} are
				recommended\footnote{\url{http://www.keylength.com/en/8/}}\textsuperscript{,}\footnote{\url{http://codahale.com/how-to-safely-store-a-password}}.
				\crit{Test coverage}{3.2} is important for this evaluation and
				for security purposes, because it lowers the probability of
				bugs per lines of code by trend. As \crit{Containability}{3.3},
				the modular independence (e.g. independence of the web
				interface from the Message Store file system) from a security
				perspective is perceived.

@@

		\subsection{The contestants}
			For introduction of "the contestants", the most important
			attributes and specialties will be listed for each. A more direct
			comparison of the features and a subsumption can be found in the
			\textsecref{sec:evaluation:overview}.

@@

				\noindent
				Of the compared tools, \emph{modoboa} seems to be one of the
				most mature ones. It uses the Python Django framework and
				therefore includes facilities like schema migrations and an
				\ac{ORM}. It passes almost every criterion; just when it comes
				to the \crit{audit trail}{2.4.1}, there is no undo support.
				Also the \ac{UI} on mobile phones is not fully elaborated and
				has some minor flaws.

@@

				\noindent
				Further information on \emph{Møil} is to be found in the
				\textsecref{sec:moeil}.

@@

				\noindent
				For many of the solutions \emph{postfix.admin} seems to be an
				inspiration and also the impulse for a new implementation. Many
				solutions aim to overcome its shortcomings, but many also fail
				to live up to that task (which in this case means every PHP
				tool). \emph{postfix.admin} uses a non-normalized \cite{dbnorm}
				database schema, which could spare a JOIN in some database
				queries but is very prone to inconsistencies. \crit{Schema
				migrations}{1.4} are not present through a database abstraction
				library or an \ac{ORM}, but manually implemented rather
				minimalist. There is one big file (\texttt{upgrade.php})
				containing a pile of functions like
				\texttt{upgrade\_<n>\_<db>}, which contain database dependent
				raw \ac{SQL} queries to perform the transformations on the
				schema. There is no \crit{systematic database
				independence}{1.5} or an \ac{ORM}, just "manual" support for
				MySQL and PostgreSQL. Oddly, for every mailbox there is an
				alias entry created (even when the mail is not forwarded).
				\emph{postfix.admin} has support for an \crit{audit
				trail}{2.4.1} but without undo. The default \crit{password
				hashing}{3.1} algorithm is \texttt{md5-crypt}. It is
				\crit{containable}{3.3} by default but does not support the
				movement of mail data, then, which would require the direct
				access to the Mail Store.

@@

				\item[\rm Responsive user interface]\ \\
					\emph{Møil} adapts to the environment it is needed in. The
					web user interface is responsive and therefore supports
					both classic desktop view ports and mobile devices like
					phones or tablets. For friends of the command line some
					commands to manage the most important resources are
					provided, too.

@@

				\item[\rm Search]\ \\
					A combined search for any resource is provided, which is
					optionally based on \ac{SQL} or
					\emph{Elasticsearch}\footnote{\url{http://www.elasticsearch.org}}
					for really huge databases (if installed and configured to
					be used).
			\end{description}

@@

			The application model is put into graphs in the following image. It
			shows the model classes and their relations among each other. The
			graph is generated, so unfortunately the actual attributes saving a
			relation and the cardinalities are missing, for now.

@@

			\subsubsection{Mailboxes}
				A \texttt{Mailbox} is connected to its \texttt{Domain} and
				possibly also to one \texttt{Relocation} and has the attributes
				\texttt{username} (which is the part in front of the
				domain in the email address), \texttt{encrypted\_password}
				(containing the password hash), \texttt{name} (for a given name
				for example), \texttt{mail\_location} (which is empty by
				default but can be used to move the users mail data location),
				\texttt{quota}, \texttt{active}, \texttt{admin} and timestamp
				attributes for saving the date of creation or the last update.
				As a password hashing algorithm, currently only
				\texttt{sha512-crypt} is supported.

			\subsubsection{Aliases}
				\texttt{Aliases} represent mail forwardings and are also
				connected to their affiliated \texttt{Domain}. They have the
				attributes \texttt{username}, \texttt{goto} (which is the
				target of the forwarding and can contain one or more internal
				or external email addresses seperated by commas),
				\texttt{active}, \texttt{description} and timestamps of
				creation and last update.

@@

			\begin{description}
				\item[\rm Better \ac{CLI}]\ \\
					Currently, Møil has seperate \ac{CLI} tools for the most
					important tasks like managing domains and mailboxes and
					importing data from an \emph{postfix.admin} installation.
					This can be expanded into a single \ac{CLI} utility with
					better support for the features the web user interface
					offers.

				\item[\rm Spam and virus check integration]\ \\
					Of the compared solutions, some integrate
					\texttt{amavisd-new} by shared database tables. After this
					being clear, there is not much left to be solved
					conceptually for a spam and virus check integration and
					this becomes a question of time.

				\item[\rm Support for server-side filters]\ \\
					The support for server-side filters in Møil was already
					examined and begun but at first a Ruby library for the
					ManageSieve protocol has to be implemented, because the
					only one existing is not maintained anymore and became
					dysfunctional.
			\end{description}
